---
title: "Assessing the quality metrics"
output: pdf_document
author: "Özlem Bahar"
date: "1/17/2022"
---


```{r packages, warning=FALSE, message = FALSE, include=FALSE}
# load the packages
require(miceadds)
require(Seurat)
require(SeuratObject)
require(ggplot2)
require(patchwork)
require(sctransform)
require(tidyverse)
require(Matrix)
require(scales)
require(cowplot)
require(RCurl)
require(readr)
require(dplyr)
```

```{r setup, include=FALSE}
# load the data
load.Rdata(file="processed_data/merged_seurat.RData", "merged_seurat")
#Create metadata dataframe
metadata <- merged_seurat@meta.data
```

## The quality control metrics

Assess the quality metrics below to identify the low quality cells to remove from the data.

* Cell counts
* UMI counts per cell
* Genes detected per cell
* UMIs vs. genes detected
* The ratio of mitochondrial mRNA count
* Novelty

```{r NCells, echo=FALSE}
# Visualize the number of cell counts per sample
metadata %>% 
  ggplot(aes(x=sample, fill=sample)) + 
  geom_bar() +
    geom_text(aes(label = ..count..),stat = "count",vjust = 1.5,colour = "white")+

  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Number of Cells")
```

## Determine the UMI counts (transcripts) per cell

The UMI count per cell should  be above 500. The UMI counts are between 500 and 1000 counts are the indication of that the cells should have been sequenced deeper.

```{r trancript/cell}
# Visualize the number UMIs/transcripts per cell
metadata %>% 
  	ggplot(aes(color=sample, x=nUMI, fill= sample)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = 500)
```

## Determine the number of genes per cell

The number of genes expected to be similar but a lower than UMIs.

For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated. If there is a small shoulder to the left of the bigger peak or a bimodal distribution, this can be indicative of couple of things:  

* Cells are distrupted.
* There are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest), and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size). 

The gene detection in the range of 500-5000 is normal for inDrop analysis. 

```{r}
# The distribution of genes detected per cell via histogram
metadata %>% 
  	ggplot(aes(color=sample, x=nGene, fill= sample)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)

# The distribution of genes detected per cell via boxplot
metadata %>% 
  	ggplot(aes(x=sample, y=log10(nGene), fill=sample)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("Number of Cells vs Number of Genes")
```

## UMIs vs. the number of genes detected

Evaluate the two metrics together: the number of UMIs and the number of genes detected per cell.The plot was colored by the fraction of mitochondrial reads. 

```{r, genes_UMIs}
# Visualize the correlation between genes detected and number of UMIs and                                         determine whether strong presence of cells with low numbers of genes/UMIs
metadata %>%
    ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) +
    geom_point() +
    scale_colour_gradient(low = "gray90", high = "black") +
    stat_smooth(method=lm) +
    scale_x_log10() +
    scale_y_log10() +
    theme_classic() +
    geom_vline(xintercept = 500) +
    geom_hline(yintercept = 500) +
    facet_wrap(~sample)
```

## The ratio of mitochondrial count

This metric is used to determine if there is a large amount of mitochondrial contamination from dead or dying cells.

```{r, mito}
# Visualize the distribution of mitochondrial gene expression detected per cell
metadata %>% 
  	ggplot(aes(color=sample, x=mitoRatio, fill=sample)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 0.25)
```

## Complexity of the RNA

Determine the complexity (novelty) of the cells by calculating the novelty score. The novelty score is calculated by taking the ratio of the number of genes over nUMI. The novelty score is expected to be above 0.80 for good quality cells.

```{r, genesperUMI}
# Determine the overall complexity of the gene expression by visualizing the                                            genes detected per UMI
metadata %>%
  	ggplot(aes(x=log10GenesPerUMI, color = sample, fill=sample)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)
```

## Filtering

# Cell-level filtering

Now that we have visualized the various metrics, we can decide on the thresholds to apply which will result in the removal of low quality cells. Often the recommendations mentioned earlier are a rough guideline, and the specific experiment needs to inform the exact thresholds chosen. We will use the following thresholds:

nUMI > 500 (This could be even 1000)
nGene > 500
log10GenesPerUMI > 0.8
mitoRatio < 0.25

```{r, vlnplot}
#Check before filtering
VlnPlot(merged_seurat, 
        features = c("nUMI", "nGene", "mitoRatio"), 
        ncol=3, 
        group.by = "sample")
```

```{r, fig.width=8, fig.height=3}
# FeatureScatter is typically used to visualize feature-feature relationships,                                          but can be used for anything calculated by the object, i.e. columns in object                                       metadata, PC scores etc.

plot1 <- FeatureScatter(merged_seurat, 
                        feature1 = "nUMI", 
                        feature2 = "mitoRatio")
plot2 <- FeatureScatter(merged_seurat, 
                        feature1 = "nUMI", 
                        feature2 = "nGene")
plot1 + plot2
```

```{r}
# Filter out low quality cells using selected thresholds 
# As the distribution of QC covariates differ between samples, QC thresholds should be determined separately for each sample (Plasschaert et al (2018))

filtered_seurat_tp1 <- subset(x = merged_seurat, subset=
                                   sample == "TP1"&
                                    mitoRatio < 0.25 & 
                                    nGene < 2500 &
                                    nGene > 500 &
                                    nUMI > 500 &
                                    log10GenesPerUMI > 0.8)

filtered_seurat_tp2 <- subset(x = merged_seurat, subset=
                                    sample == "TP2"&
                                    mitoRatio < 0.25 &
                                   nGene > 500 &
                                     nUMI > 500 &
                                     log10GenesPerUMI > 0.8)

filtered_seurat_tp2.2 <- subset(x = merged_seurat, subset=
                                   sample == "TP2.2"&
                                    mitoRatio < 0.1 &
                                    nGene < 4500 &
                                    nGene > 500 &
                                    nUMI > 500 &
                                    log10GenesPerUMI > 0.8)

filtered_seurat <- merge(x = filtered_seurat_tp1,
                          y = c(filtered_seurat_tp2,
                                filtered_seurat_tp2.2))

```

## Gene-level filtering

Genes with zero counts dramatically reduce the average expression for a cell and therefore should be removed.

```{r}
# Identify genes with zero counts
object = filtered_seurat
counts <- GetAssayData(object = filtered_seurat, slot = "counts")

# Output a logical matrix specifying for each gene on whether or not there are                                          more than zero counts per cell
nonzero <- counts > 0
```

Keep the genes which are expressed in 10 or more cells and thereby remove the genes with zero counts in all cells.

```{r}

# returns TRUE if more than 10 TRUE values per gene
keep_genes <- Matrix::rowSums(nonzero) >= 10

# keep those expressed in more than 10 cells
filtered_counts <- counts[keep_genes, ]
```

Generate a new Seurat object with the remaning cells for downstream analysis.

```{r}
# Reassign to filtered Seurat object
filtered_seurat <- CreateSeuratObject(filtered_counts, 
                                      meta.data = filtered_seurat@meta.data)
```

```{r}
#Check the metrics after filtering
VlnPlot(filtered_seurat, features = c("nUMI", "nGene", "mitoRatio"), ncol=3)
```

```{r}
filtered_seurat@meta.data -> metadata
```


```{r}
metadata %>% 
  ggplot(aes(x=sample, fill=sample)) + 
  geom_bar()+
  geom_text(aes(label = ..count..),stat = "count",vjust = 1.5,colour = "white")+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Number of Cells")
```


```{r}
# Save filtered object
save(filtered_seurat, file="processed_data/filtered_seurat.RData")
```
