---
title: "Clustering the cells and marker identification"
output: pdf_document
author: "Bahar,Y."
date: "4/26/2023"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
require(Seurat)
require(tidyverse)
require(RCurl)
require(cowplot)
require(miceadds)
```


```{r rdata}
#load the .RDS object from the 2nd step 
seurat_integrated <- readRDS(file="processed_data/integrated_seurat.rds")
```


## Cluster the cells based on top PCs 
```{r pc.heatmap}
# Explore heatmap of PCs
DimHeatmap(seurat_integrated, 
           dims = 1:15, 
           cells = 500, 
           balanced = TRUE)
```

Print out the top 10 (or more) positive and negative genes by PCA scores driving the PCs.

```{r variable.genes}
# Check the most variable genes driving PCs
print(x = seurat_integrated[["pca"]], 
      dims = 1:15, 
      nfeatures = 20)
```


# Plot the elbow plot to visualize the standard deviation of each PC and to determine how many PCs to use for clustering (where the standard deviations begins to plateau)

```{r}
# Plot the elbow plot
ElbowPlot(object = seurat_integrated, 
          ndims = 50)
```
## Cluster the cells

```{r, k.near, include=FALSE}
# Determine the K-nearest neighbor graph
seurat_integrated <- FindNeighbors(object = seurat_integrated, 
                                dims = 1:50)
                                
# Determine the clusters for various resolutions
# separate column for metadata in seurat_integrated@meta.data
seurat_integrated <- FindClusters(object = seurat_integrated,
                               resolution = c(0.6, 0.8))
```


```{r}
# Visualize the different resolutions
seurat_integrated@meta.data %>% 
        View()
```

```{r, fig.height=10, fig.width=10}
# Assign identity of clusters
Idents(object = seurat_integrated) <- "integrated_snn_res.0.6"

# Plot the UMAP
DimPlot(seurat_integrated,
        reduction = "umap",
        label = TRUE,
        label.size = 6)
```

## Check the quality control metrics

```{r}
# Create sample column
seurat_integrated@meta.data$orig.ident[which(str_detect(seurat_integrated@meta.data$cells, "^TP1_"))] <- "tp1"
#seurat_integrated@meta.data$orig.ident[which(str_detect(seurat_integrated@meta.data$cells, "^TP2_"))] <- "tp2"
seurat_integrated@meta.data$orig.ident[which(str_detect(seurat_integrated@meta.data$cells, "^TP2.2_"))] <- "tp2.2"
```


```{r}
# Determine the number of cells per cluster per sample
n_cells <- FetchData(seurat_integrated, 
                     vars = c("ident", "orig.ident")) %>%
        dplyr::count(ident, orig.ident) %>%
        tidyr::spread(ident, n)

View(n_cells)
```

Check the nummber of the cells per cluster for each sample using the UMAP:

```{r}
# UMAP of cells in each cluster by sample
DimPlot(seurat_integrated, 
        label = TRUE, 
        split.by = "sample")  + NoLegend()
```

# Check for segregation of clusters by S-phase and G2M-phase markers

```{r}
# Plot the metrics and check if they vary the clusters
metrics <-  c("nCount_RNA", "nFeature_RNA", "mitoRatio")

FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = metrics,
            pt.size = 0.4, 
            sort.cell = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

## Which PCs do drive the different clusters?

PCs should separate the clusters.
```{r}
# Get the UMAP info
columns <- c(paste0("PC_", 1:50),
            "ident",
            "UMAP_1", "UMAP_2")

# Extract the data
pc_data <- FetchData(seurat_integrated, 
                     vars = columns)
```


```{r}
# Extract the UMAP coordinates for the first 10 cells
seurat_integrated@reductions$umap@cell.embeddings[1:10, 1:2]
```

```{r}
# Add the cluster name
umap_label <- FetchData(seurat_integrated, 
                        vars = c("ident", "UMAP_1", "UMAP_2"))  %>%
  group_by(ident) %>%
  summarise(x=mean(UMAP_1), y=mean(UMAP_2))
  
#Plot the first 16 PCs
map(paste0("PC_", 1:15), function(pc){
        ggplot(pc_data, 
               aes(UMAP_1, UMAP_2)) +
                geom_point(aes_string(color=pc), 
                           alpha = 0.7) +
                scale_color_gradient(guide = FALSE, 
                                     low = "grey90", 
                                     high = "blue")  +
                geom_text(data=umap_label, 
                          aes(label=ident, x, y)) +
                ggtitle(pc)
}) %>% 
        plot_grid(plotlist = .)
```


```{r}
# Check the PCA results 
print(seurat_integrated[["pca"]],  dims = 1:15, nfeatures = 5)
```

```{r}
# Select the RNA counts slot to be the default assay
DefaultAssay(seurat_integrated) <- "RNA"

# Normalize RNA data for visualization purposes
seurat_integrated <- NormalizeData(seurat_integrated, verbose = FALSE)
```

## Marker identification

```{r, conserved.markers}
# require(multtest)
# require(metap)
FindConservedMarkers(seurat_integrated,
                     ident.1 = 0,
                     grouping.var = "sample",
                     only.pos = TRUE,
		                 min.diff.pct = 0.25,
                     min.pct = 0.25,
		                 logfc.threshold = 0.25)
```

```{r}
cluster0_conserved_markers <- FindConservedMarkers(seurat_integrated,
                              ident.1 = 0,
                     	        grouping.var = "sample",
                              only.pos = TRUE,
		                          logfc.threshold = 0.25)
```

```{r}
cluster12_conserved_markers <- FindConservedMarkers(seurat_integrated,
                              ident.1 = 12,
                     	        grouping.var = "sample",
                              only.pos = TRUE,
		                          logfc.threshold = 0.25)
```

## Adding the gene annonation

```{r}
#Get the gene annotation information from annotation.csv
annotations <- read.csv(file="processed_data/annotations.csv")
```


```{r}
# Merge the gene annonation info. 
cluster0_ann_markers <- cluster0_conserved_markers %>% 
                rownames_to_column(var="gene") %>% 
                left_join(y = unique(annotations[, c("gene_name", "description")]),
                          by = c("gene" = "gene_name"))

View(cluster0_ann_markers)
```


```{r, markers.by.cluster}
# A function to identify markers for any given cluster
get_conserved <- function(cluster){
  FindConservedMarkers(seurat_integrated,
                       ident.1 = cluster,
                       grouping.var = "sample",
                       only.pos = TRUE) %>%
    rownames_to_column(var = "gene") %>%
    left_join(y = unique(annotations[, c("gene_name", "description")]),
               by = c("gene" = "gene_name")) %>%
    cbind(cluster_id = cluster, .)
  }
```

```{r}
# Iterate function across desired clusters
conserved_markers <- map_dfr(c(0:23), get_conserved)
```
## Evaluating marker genes

Check for top 50 markers to identify the clusters
```{r}
# Extract top 50 markers per cluster
top50_merged <- conserved_markers %>% 
  mutate(avg_fc = (TP1_avg_log2FC + TP2.2_avg_log2FC) /2) %>% 
  group_by(cluster_id) %>% 
  top_n(n = 50, 
        wt = avg_fc)

# Visualize top markers per cluster
#View(top50)
library(openxlsx)
#write.xlsx(top50_merged, file = "markers/top50_merged.xlsx", colNames = TRUE, rownames = TRUE)
```

```{r}
# Rename all identities
seurat_integrated_labelled <- RenameIdents(object = seurat_integrated, 
                               "0" = "VNC primordium",
                               "1" = "Newborn neuron",
                               "2" = "NB",
                               "3" = "NB",
                               "4" = "GMC",
                               "5" = "NB",
                               "6" = "NB",
                               "7" = "Neural precursor",
                               "8" = "NB",
                               "9" = "NB",
                               "10" = "VNC primordium",
                               "11" = "NB",
                               "12"= "Glia",
                               "15" = "Glia precursor",
                               "13" = "Unknown",
                               "14" = "Unknown",
                               "16" = "Midline primordium",
                               "17" ="Neural precursor", 
                               "18" = "NB",
                               "19" ="VNC primordium",
                               "20"="Sensory complex primordium",
                               "21" ="Unknown",
                               "22"= "Neural precursor",
                               "23" ="Unknown")

# Plot the UMAP
DimPlot(object = seurat_integrated_labelled,
        reduction = "umap", 
        label = TRUE,
        label.size = 6,
        repel = TRUE)+
        theme_void()+
        theme(legend.position = "none")

#ggsave("Figures/step1.png", width=6, height=5, dpi=400)
```

```{r}
n_cells_clusters <- FetchData(seurat_integrated_labelled, 
                     vars = c("ident", "orig.ident")) %>%
        dplyr::count(ident, orig.ident) %>%
        tidyr::spread(ident, n)
```

```{r}
# Save the final R object
write_rds(seurat_integrated,
          file = "processed_data/4_seurat_integrated_labelled.rds")

# Create and save a text file with sessionInfo
# sink("sessionInfo_scrnaseq_March23.txt")
# sessionInfo()
# sink()
```