---
title: "Dissecting individual NB clusters"
author: "Bahar, Y."
date: "4/26/2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
require(Seurat)
require(tidyverse)
require(RCurl)
require(cowplot)
require(miceadds)
```

```{r}
seurat_integrated_labelled <- readRDS(file = "processed_data/4_seurat_integrated_labelled.rds")
```

```{r fig.height=6, fig.width=8}
# Plot the UMAP
 DimPlot(object = seurat_integrated_labelled, 
        reduction = "umap", 
        label = TRUE,
        label.size = 6,
        repel = TRUE)
```

```{r}
# Check how many cells we have
seurat_integrated_labelled
```


```{r}
# Extract UMAP coordinates from the Seurat object
umap_coords <- seurat_integrated_labelled@reductions$umap@cell.embeddings

# Save UMAP coordinates to a CSV file
write.csv(umap_coords, "UMAP_style/umap_coords_dseurat_integrated_labelled.csv", row.names = FALSE)
```

# Approach 2 chosen: SCT reintegration > Re-cluster 
(following Integration tutorial)

This approach was used for the rest of the analyses as it provided better
results in the clustering of cells.
```{r}
#Get DEGs between NBs and all other
nb.dges <- FindMarkers(seurat_integrated_labelled, ident.1 = "NB", ident.2 = NULL, only.pos = TRUE)
# view results
head(nb.dges)
write.xlsx(nb.dges, file = "markers/nb.dges.up.xlsx", colNames = TRUE, rownames = TRUE)
```

## Plot dpn+ cells

```{r}
FeaturePlot(seurat_integrated_labelled, 
            reduction = "umap", 
            features = c("dpn"), 
            min.cutoff = 'q10') +labs(title='NBs (dpn)')
#ggsave("figures/interesting_vn_cluster_sep/dpn_pos_alltps.png", width=8, height=7)
```

```{r dpn+, include=FALSE, warning=FALSE}
#Subseting
DefaultAssay(seurat_integrated_labelled) <- "RNA"
dpn.seurat<- subset(seurat_integrated_labelled, subset=dpn>1)
#
dpn.seurat_list2 <- SplitObject(dpn.seurat, split.by = "sample")
dpn.seurat_list2 <- lapply(dpn.seurat_list2, SCTransform, vars.to.regress = "mitoRatio")
features  <- SelectIntegrationFeatures(dpn.seurat_list2, nfeatures = 2000)
dpn.seurat_list2 <- PrepSCTIntegration(dpn.seurat_list2, anchor.features = features)
anchors <- FindIntegrationAnchors(dpn.seurat_list2, normalization.method = "SCT", anchor.features = features)
dpn.seurat_combined2 <- IntegrateData(anchorset = anchors, normalization.method = "SCT", k.weight = 46)
dpn.seurat_combined2 <- RunPCA(dpn.seurat_combined2) 
dpn.seurat_combined2 <- RunUMAP(dpn.seurat_combined2, reduction = "pca", dims = 1:50)
dpn.seurat_combined2 <- FindNeighbors(dpn.seurat_combined2, dims = 1:50)
dpn.seurat_combined2 <- FindClusters(dpn.seurat_combined2, dims = 1:50, resolution = c(0.6, 0.8))
#ElbowPlot(object = dpn.seurat_combined2,ndims = 40)
#write_rds(dpn.seurat_combined2, file = "processed_data/dpn.seurat_combined2.rds")
```

```{r}
# Check how many cells left after selecting dpn expressing cells
#dpn.seurat_combined2 <-  readRDS(file = "processed_data/dpn.seurat_combined2.rds")
dpn.seurat_combined2
```

```{r fig.height=6, fig.width=8}
# Assign identity of clusters
Idents(object = dpn.seurat_combined2) <- "integrated_snn_res.0.8"

# Plot the UMAP
DimPlot(dpn.seurat_combined2,
        reduction = "umap",
        label = TRUE,
        label.size = 4,
        repel=TRUE)

#ggsave("figures/interesting_vn_cluster_sep/dpn_subclusters.png", width=8, height=7)
```

```{r}
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "fne", 
            min.cutoff = 'q10') +
            ggtitle("Neuron (fne)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
#ggsave("/Figures/Step2_fne.png", width=6, height=5, dpi=400)

FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "tap", 
            min.cutoff = 'q10') +
            ggtitle("GMCs (tap)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
#ggsave("/Figures/Step2_tap.png", width=6, height=5, dpi=400)
```

```{r}
#Extract top markers to filter out populations of cells other than NBs
dpn.seurat_combined2.markers = FindAllMarkers(dpn.seurat_combined2, 
                                              only.pos = TRUE, 
                                              min.pct = 0.25, 
                                              logfc.threshold = 0.25)

#Extract top50 markers
# #top50_dpn.pos <- dpn.seurat_combined2.markers %>% 
#   mutate(avg_fc = avg_log2FC) %>% 
#   group_by(cluster) %>% 
#   top_n(n = 50, 
#   wt = avg_fc)
```
#Annotate dpn+ clusters
```{r fig.height=6, fig.width=8}
dpn.seurat_combined2_annotated<- RenameIdents(object =dpn.seurat_combined2,
"12"="Neuron",
"15" ="Newborn neuron",
"3" = "Newborn neuron",
"0" = "GMC",
"1" = "VNC",
"2" = "VNC",
"4" = "VNC",
"5" = "VNC",
"6"="VNC",
"7"="VNC/Brain primordium",
"8" = "VNC/Mesoderm primordium",
"9"= "VNC",
"10"= "VNC/Brain primordium",
"11"="VNC/Brain primordium",
"13"= "VNC",
"14"="VNC",
"16" = "NGB/GB",
"17" = "VNC",
"18"="VNC",
"19"= "Unknown",
"20"= "Unknown")

# Plot the UMAP
DimPlot(object = dpn.seurat_combined2_annotated, 
        reduction = "umap", 
        label = TRUE,
        label.size = 6,
        repel = TRUE)+
        theme_void()+
        theme(legend.position = "none")



#ggsave("/Figures/step2.png", width=6, height=5, dpi=400)
```

```{r}
# Generate the heatmap for step2
dpn.seurat_combined2_annotated_markers = FindAllMarkers(dpn.seurat_combined2_annotated, 
                                              only.pos = TRUE, 
                                              min.pct = 0.25, 
                                              logfc.threshold = 0.25)
dpn.seurat_combined2_annotated_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10


DoHeatmap(dpn.seurat_combined2_annotated, features = top10$gene)
  
ggsave("/Figures/heatmap_step2.png", height= 10, width=10, dpi=400)
```

```{r}
# Extract UMAP coordinates from the Seurat object
umap_coords <- dpn.seurat_combined2_annotated@reductions$umap@cell.embeddings

# Save UMAP coordinates to a CSV file
write.csv(umap_coords, "UMAP_style/umap_coords_dpn.seurat_combined2_annotated.csv", row.names = FALSE)
```

#Explore the clusters of dpn+ cells
```{r}
require(dplyr)
# Retrieve the top 5 marker genes per cluster
# Use whichever genes have the highest values under the AVG_LOG column
top5 <- dpn.seurat_combined2.markers %>% group_by(cluster) %>%
  dplyr::slice_max(get(grep("^avg_log", colnames(dpn.seurat_combined2.markers), value = TRUE)),
                   n = 5)
# Create the dot plot
Seurat::DotPlot(dpn.seurat_combined2_annotated, features = unique(top5$gene)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 1,
                                            size = 12, hjust = 1)) +
  Seurat::NoLegend()

#ggsave("figures/step2_dpn_poss_clusters.svg", width=20, height=10)
# Create the heatmap
Seurat::DoHeatmap(dpn.seurat_combined2_annotated, features = unique(top5$gene)) +
  Seurat::NoLegend() +
  ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))
```


Note: Some of the dpn positive cells, show a variety of different cell type characteristics. This might be indication of they started to specialize. These more specialized cells should be filtered out from the data.
```{r}
#cluster 12: fne is a neuron marker
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "fne", 
            min.cutoff = 'q10') +
            ggtitle("Neuron (fne)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
ggsave("/Figures/Step2_fne.png", width=6, height=5, dpi=400)
#cluster3: E(spl)m4-BFM is a neural progenitor marker
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "E(spl)m4-BFM", 
            min.cutoff = 'q10') +
            ggtitle("Newborn neuron (E(spl)mγ-HLH)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "gcm", 
            min.cutoff = 'q10') +
            ggtitle("NGB/GB(gcm)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "Traf1", 
            min.cutoff = 'q10') +
            ggtitle("GMC (Traf1)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
FeaturePlot(dpn.seurat_combined2, 
            reduction = "umap", 
            features = "sna", 
            min.cutoff = 'q10') +
            ggtitle("VNC(sna)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
ggsave("/Figures/Step2_sna.png", width=6, height=5, dpi=400)
```

## Get the clusters expressing domain markers 

```{r, warning=FALSE ,include=FALSE}

DefaultAssay(dpn.seurat_combined2) <- "RNA"
#subset VNC populations
dpn.vnc.domain<- subset(dpn.seurat_combined2, idents=c("1", "2", "4", "5", "6", "7", "8","9","10","11","13","14","16", "17", "18"))

dpn.vnc.domain_list <- SplitObject(dpn.vnc.domain, split.by = "sample")
dpn.vnc.domain_list <- lapply(dpn.vnc.domain_list, SCTransform, vars.to.regress = "mitoRatio")
features  <- SelectIntegrationFeatures(dpn.vnc.domain_list, nfeatures = 2000)
dpn.vnc.domain_list <- PrepSCTIntegration(dpn.vnc.domain_list, anchor.features = features)
anchors <- FindIntegrationAnchors(dpn.vnc.domain_list, normalization.method = "SCT", anchor.features = features)
dpn.vnc.domain_list<- IntegrateData(anchorset = anchors, normalization.method = "SCT", k.weight = 46)
dpn.vnc.domain_list <- RunPCA(dpn.vnc.domain_list) 
dpn.vnc.domain_list <- RunUMAP(dpn.vnc.domain_list, reduction = "pca", dims = 1:20)
dpn.vnc.domain_list <- FindNeighbors(dpn.vnc.domain_list, dims = 1:20)
dpn.vnc.domain_list <- FindClusters(dpn.vnc.domain_list, dims = 1:20, resolution = c(0.6, 0.8))
ElbowPlot(object = dpn.vnc.domain_list, ndims = 40)
```

```{r}
# Check how many cells left and save the .R object
#dpn.vnc.domain_list
#write_rds(dpn.vnc.domain_list, file = "processed_data/dpn.vnc.domain_list.rds")
```

```{r}
# Plot the UMAP
# Assign the identity of clusters

Idents(object = dpn.vnc.domain_list) <- "integrated_snn_res.0.8"

# Plot the UMAP
DimPlot(dpn.vnc.domain_list,
        reduction = "umap",
        label = TRUE,
        label.size = 6,
        repel=TRUE)

#ggsave("figures/interesting_vn_cluster_sep/dpn_vnc_subcluster.png", width=8, height=7)
```

```{r}
# Extract UMAP coordinates from the Seurat object
umap_coords <- dpn.vnc.domain_list@reductions$umap@cell.embeddings

# Save UMAP coordinates to a CSV file
write.csv(umap_coords, "UMAP_style/umap_coordsdpn.vnc.domain_list.csv", row.names = FALSE)
```

#Dissect NBs from midline, brain and ventral neuroectoderm.

```{r}
#Extract top 50 markers
#2nd filter: Identified neuroectoderm drived NBs and filter out midline and brain drived NBs
dpn.vnc.domain_list.markers = FindAllMarkers(dpn.vnc.domain_list, 
                                              only.pos = TRUE, 
                                              min.pct = 0.25, 
                                              logfc.threshold = 0.25)

#Extract top 50 markers
top50.nenbs <- dpn.vnc.domain_list.markers %>% 
  mutate(avg_fc = avg_log2FC) %>% 
  group_by(cluster) %>% 
  top_n(n = 50, 
        wt = avg_fc)

# Visualize top 10 markers per cluster
View(top50.nenbs)
#Save the top 50
#write.xlsx(top50.nenbs, file = "markers/3-top50.nb.dissected.xlsx", colNames = TRUE, rownames = TRUE)

```

```{r fig.height=6, fig.width=8}
dpn.vnc.domain_list_annotated<- RenameIdents(object =dpn.vnc.domain_list,
"0"="NB",
"1" ="NB",
"2" = "NB",
"3" = "NB",
"4" = "NB",
"5" = "MG",
"6" = "NB",
"7" = "Brain/Midline NB",
"8"="Brain NB",
"9"="NB",
"10" = "NB",
"11"= "Brain/Midline NB",
"12"= "NB",
"13"="NB",
"14"= "NB",
"14"= "NB",
"15"= "NB")


# Plot the UMAP
# DimPlot(object = dpn.vnc.domain_list_annotated, 
#         reduction = "umap", 
#         label = TRUE,
#         label.size = 6,
#         repel = TRUE)+
#         theme_void()+
#         theme(legend.position = "none")

#ggsave("/Figures/step3_labelled.png", width=6, height=5, dpi = 400)
```


```{r}
#Generate the heatmap for step 3
dpn.vnc.domain_list_annotated_markers = FindAllMarkers(dpn.vnc.domain_list_annotated, 
                                              only.pos = TRUE, 
                                              min.pct = 0.25, 
                                              logfc.threshold = 0.25)
dpn.vnc.domain_list_annotated_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10


DoHeatmap(dpn.vnc.domain_list_annotated, features = top10$gene)
  
ggsave("/Figures/heatmap_step3.png", height= 10, width=8, dpi=400)
```


```{r MidlineBrainNBs}

FeaturePlot(dpn.vnc.domain_list_annotated, 
            reduction = "umap", 
            features = "cas", 
            min.cutoff = 'q10') +
            ggtitle("Brain+Midline NB (cas)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
ggsave("/step3_BrainMidlineNB.png", width=6, height=5, dpi = 400)

FeaturePlot(dpn.vnc.domain_list_annotated, 
            reduction = "umap", 
            features = "toy", 
            min.cutoff = 'q10') +
            ggtitle("Brain NB (toy)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
#plot tll expression and save
FeaturePlot(dpn.vnc.domain_list_annotated, 
            reduction = "umap", 
            features = "tll", 
            min.cutoff = 'q10') +
            ggtitle("Brain NB (tll)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))
ggsave("/Figures/step3_BrainNB_tll.png", width=6, height=5, dpi = 400)
#plot oc expression
FeaturePlot(dpn.vnc.domain_list_annotated, 
            reduction = "umap", 
            features = "oc", 
            min.cutoff = 'q10') +
            ggtitle("Brain/Midline NB (oc)")+
            theme_void()+
            theme(plot.title = element_text(hjust = 0.5, vjust = 1, size=20, face="bold"))

ggsave("/Users/Figures/step3_BrainNB.png", width=6, height=5, dpi = 400)

#Notes:
#spi is expressed ubiquitously and other midline specific markers are not marking         #any clusters. So my strategy to identify midline NBs as follows:
#vvl and cas are markers for both midline and brain. Since we identified brain          #NBs with brain specific marker toy, the remaining cluster marked via vvl and       
#cas labels midline. (Only early time points of development)
#ggsave("figures/midline.jpeg", width=15, height=10)
```

# Take only the VNC NB clusters as well as Midline Glia (MG)

```{r, warning=FALSE ,include=FALSE}
# Subcluster after removing brain/midline NBs

DefaultAssay(dpn.vnc.domain_list) <- "RNA"
#subset AP neuroblast populations
dpn.nb.clusters<- subset(dpn.vnc.domain_list, idents = c("0", "1", "2", "3", "4", "5", "6", "9", "10", "12", "13", "14", "15"))

dpn.nb.clusters_list <- SplitObject(dpn.nb.clusters, split.by = "sample")
dpn.nb.clusters_list <- lapply(dpn.nb.clusters_list, SCTransform, vars.to.regress = "mitoRatio")
features  <- SelectIntegrationFeatures(dpn.nb.clusters_list, nfeatures = 2000)
dpn.nb.clusters_list <- PrepSCTIntegration(dpn.nb.clusters_list, anchor.features = features)
anchors <- FindIntegrationAnchors(dpn.nb.clusters_list, normalization.method = "SCT", anchor.features = features)
dpn.nb.clusters_list<- IntegrateData(anchorset = anchors, normalization.method = "SCT",k.weight = 46)
dpn.nb.clusters_list <- RunPCA(dpn.nb.clusters_list) 
dpn.nb.clusters_list <- RunUMAP(dpn.nb.clusters_list, reduction = "pca", dims = 1:15)
dpn.nb.clusters_list <- FindNeighbors(dpn.nb.clusters_list, dims = 1:20)
dpn.nb.clusters_list <- FindClusters(dpn.nb.clusters_list, dims = 1:20, resolution = c(0.6, 0.8, 1))
ElbowPlot(object = dpn.nb.clusters_list, ndims = 40)
```

```{r fig.height=6, fig.width=8}
# Plot the UMAP
# Assign identity of clusters
Idents(object = dpn.nb.clusters_list) <- "integrated_snn_res.0.8"

# Plot the UMAP
DimPlot(dpn.nb.clusters_list,
        reduction = "umap",
        label = TRUE,
        label.size = 6,
        repel=TRUE)

#ggsave("figures/interesting_vn_cluster_sep/dpn.nb.clusters_list.png", width=8, height=7)
```

```{r fig.height=6, fig.width=8}
DotPlot(object = dpn.nb.clusters_list, 
                         features = c("Dr","ind","vnd", "en","wg","gsb", "svp", "ac","hkb", "slp1","lsc","odd", "mirr", "CG42342" ,"retn","pnt", "htl", "gcm"), 
                         assay="RNA")+
       theme(axis.text.x = element_text(angle = 45))
#ggsave("figures/interesting_vn_cluster_sep/dotplot_dpn.nb.clusters_list.png", width=8, height=7)
```

```{r}
#Check the expression of Fer3 or Fer1 (enriched in the data)
FeaturePlot(dpn.nb.clusters_list, 
            reduction ="umap", 
            features = "Fer1", 
            label = TRUE,
            min.cutoff = 'q10') +labs(title='Fer1')
#ggsave("figures/fixed-result-apr/3-nb.clu/Fer1.png", width=9, height=7)
```

```{r}
#Save before labeling
#write_rds(dpn.nb.clusters_list, file = "processed_data/dpn.nb.clusters_list.rds")
dpn.nb.clusters_list <- readRDS(file = "processed_data/dpn.nb.clusters_list.rds")
```

```{r, fig.height=7, fig.width=10}
dpn.nb.clusters_list<- RenameIdents(object =dpn.nb.clusters_list,
"5" = "7-4",
"4" = "7-4",
"8" = "7-1",
"13" = "5-6",
"0" = "5-3",
"14" = "5-2",
"6"="6-2",
"12"="5-4",
"11" = "4-2",
"15"= "3-5",
"9"= "3-2",
"10"="2-5",
"7"="2-4",
"1"= "1-2",
"3"="1-1",
"2" ="MG")

# Plot the UMAP
DimPlot(object = dpn.nb.clusters_list, 
        reduction = "umap", 
        label = TRUE,
        label.size = 6,
        repel = TRUE)+
        theme_void()+
        theme(legend.position = "none")

#ggsave("/Figures/step4_NBs.png", width=6, height=5, dpi=400)
```

```{r}
# Extract UMAP coordinates from the Seurat object
umap_coords <- dpn.nb.clusters_list@reductions$umap@cell.embeddings

# Save UMAP coordinates to a CSV file
write.csv(umap_coords, "UMAP_style/umap_coords_dpn.nb.clusters_list.csv", row.names = FALSE)
```


```{r fig.height=6, fig.width=8}
DotPlot(object = dpn.nb.clusters_list, features = c("Dr","ind","vnd", "en","wg","gsb", "svp", "ac","hkb","odd", "mirr", "CG42342" ,"retn","pnt", "htl", "run", "gcm", "sim","eg", "Poxn"), 
                           assay="RNA")+
       theme(axis.text.x = element_text(angle = 45))
#ggsave("Figures/dotplot-nbs.png", width=9 , height = 6)
```

```{r}
# The heatmap for given cells and features
dpn.nb.clusters_list_markers = FindAllMarkers(dpn.nb.clusters_list, 
                                              only.pos = TRUE, 
                                              min.pct = 0.25, 
                                              logfc.threshold = 0.25)
dpn.nb.clusters_list_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10


#DoHeatmap(dpn.nb.clusters_list, features = top10$gene)+
  theme(
    axis.text.x = element_text(size = 25), 
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14)
  )
#ggsave("/Figures/heatmap_nbs.png", height=18, width=12, dpi=600)
```

```{r}
# Save the final R object
write_rds(dpn.nb.clusters_list, file = ("processed_data/dpn.nb.clusters_list.rds"))
#Create and save a text file with sessionInfo
#sink("sessionInfo_scrnaseq_Apr25.txt")
#sessionInfo()
#sink()
```
